Correlation tests

### 2025-05-07 analysis/manuscript.tex
Added two important paper and submitted to journals

These discussions highlight the need for robust data to evaluate
transformative agreements. Importantly, these agreements themselves can
help improve this evidence base, as researchers and libraries advocate
for including open metadata in negotiations to help avoid data analytics
from becoming a commodity controlled by publishers (Aspesi \& Brand,
2020; McCabe \& Mueller-Langer, 2024). For instance, ESAC recommends
using Crossref to share metadata supporting workflows including open
access licenses and funding information to support discovery and
monitoring (Geschuhn \& Stone, 2017). Simultaneously, the bibliometric
community pushes for more open metadata through this publisher-driven
DOI registration platform and continuously monitors progress (van Eck \&
Waltman, 2024).

@@

While promising for general bibliometric research, applying open
metadata to transformative agreements presents distinct challenges due
to their targeted journal coverage and complex funding attribution. With
invoicing data largely unavailable, previous studies have relied on
publication fee pricing lists or expenditures, and first and/or
corresponding author affiliations as proxies to attribute funding to
institutions (Butler et al., 2023; Haustein et al., 2024; Zhang et al.,
2022). Recent work has extended this approach and attributed open access
articles to transformative agreements by using data from the cOAlition S
Journal Checker Tool, which links journals and participating
institutions to transformative agreements included in the ESAC registry
(de Jonge et al., 2025; Jahn, 2025). Yet, no systematic comparison
exists of how open versus proprietary data sources perform when
analysing the same hybrid journals covered by transformative agreements
to identify comparative strengths and shortcomings of data sources.

@@

hoaddata, developed and maintained by the author to support open access
monitoring and research (Achterberg \& Jahn, 2023; Jahn, 2025), is an R
data package that regularly collects information on hybrid open access
uptake from multiple openly available data sources (Jahn, 2024). It
combines article-level metadata from Crossref (Hendricks et al., 2020)
and affilation metadata from OpenAlex (Priem et al., 2022) with
transformative agreement information from the Transformative Agreement
Data dump used by the cOAlition S Journal Checker Tool (JCT)\footnote{\url{https://journalcheckertool.org/transformative-agreements/}},
which links journal and institutional data about participating research
organisations to agreements in the ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}.

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
JCT journal data were enriched with ISSN variants linked to ISSN-L
according to the ISSN agency. To identify hybrid journals, a
comprehensive exclusion of fully open access journals was performed
using multiple journal lists including the Directory of Open Access
Journals (DOAJ). The resulting hybrid journal data comprising more tahn
13,000 hybrid journals were made available via hoaddata and were used to
independently determine the publication volume for each database.

@@

Figure \ref{fig-upset_coverage_results} presents the coverage of hybrid
journals included in transformative agreements, based on the
intersections of the examined data sources. The intersection sets of
journals (\ref{fig-upset_coverage_results}A) and articles
(\ref{fig-upset_coverage_results}B= are visualised as an UpSet graph
(Krassowski, 2020; Lex et al., 2014). The x-axis represents the set
intersections in a matrix layout. The analysis included hybrid journals
that published at least one open access article between 2019 and 2023,
based on open access status information from each database. Only
original articles were included in the analysis.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 focusing on publisher portfolios. The analysis highlights
the dominance of the three largest publishers, Elsevier, Springer
Nature, and Wiley, which collectively accounted for 47\% of the hybrid
journals and 62\% of the articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in 2,311
hybrid journals, although it recorded the largest number of hybrid
journals (19\%). Wiley accounted for 858,939 articles (12\%) in 1,382
hybrid journals (11\%). The remaining 54 publishers collectively
accounted for 2,749,847 articles (38\%) in 6,476 hybrid journals (53\%).

@@

Similarly, hybrid open access by transformative agreements substantially
increased between 2019 and 2023 (Figures \ref{fig-uptake_overview}B and
C). Trends were consistent when measuring the first (dashed line) and
corresponding authorship (dotted line). According to Scopus, 479,297
open access articles could be attributed to transformative agreements
based on first author metadata (increasing from 15,341 to 206,084) and
489,262 using corresponding author metadata (from 15,444 to 210,816).
Web of Science recorded 493,028 open access articles via transformative
agreements using first author metadata (increasing from 21,871 to
174,126) and 500,076 using corresponding author metadata (from 22,092 to
177,030). hoaddata, lacking corresponding author data, linked 501,649
articles to transformative agreements, increasing from 16,010 in 2019 to
183,757 in 2023.

@@

When examining the percentage of open access articles in hybrid journals
(Figure \ref{fig-uptake_country}B), a different pattern emerged. Authors
affiliated with institutions from medium-sized European countries, such
as Sweden, the Netherlands, Finland, and Hungary, made a large
proportion of their articles open access. Germany and the United Kingdom
also had approximately 40\% of their output available as open access. In
contrast, non-OECD countries showed notably lower adoption of hybrid
open access, with South Africa being the only BRICS member (as of 2022)
well-represented in the data. The United States also demonstrated a
relatively low proportion of open access articles. These findings were
consistent across all databases. However, France was better represented
in Web of Science, likely because of its agreement with Elsevier
starting in 2019, which allowed delayed open access under the
publisher's user license (Rabesandratana, 2019). This licence was not
classified as hybrid open access in either Scopus or hoaddata. In all
cases, Spearman rank correlations were \(\rho > .9\), \(p < 0.001\),
indicating a high level of correlation between the databases and
authorship roles considered.

@@

The coverage analysis revealed that hybrid journals are well indexed in
all three data sources, particularly in terms of article coverage, which
reflects the dominance of major publishers whose established journal
portfolios are comprehensively indexed in proprietary databases (Bellen
et al., 2024). Differences emerge for journals targeting practitioners
or local non-English language communities, with many such titles indexed
exclusively in Crossref and Scopus. Using Crossref as a bibliometric
database in hoaddata demonstrated a particular strength in identifying
newly established hybrid journals, a notable finding given that
transformative agreements primarily target existing subscription-based
journals. The landscape of hybrid journal publishing thus differs
markedly from that of fully open access journals. Comparing the coverage
of OpenAlex, Scopus and Web of Science, Simard et al. (2024) indicate
that only half of the fully open access journals listed in the DOAJ are
also indexed in Scopus and Web of Science. Notably, journals that charge
no publication fees (``diamond journals'') are absent from the selective
Web of Science, which reinforces existing disparities in the indexing of
under-represented research communities and regions in selective
bibliometric databases (Simard et al., 2024).

@@

The database comparison revealed important discrepancies in open access
evidence. Crucially, not all publishers share CC licence metadata via
Crossref, a limitation that becomes apparent when contrasting Crossref
with Unpaywall's data in Scopus and Web of Science. While Unpaywall can
detect such gaps by parsing publishers' websites for open access
licences, it nevertheless missed a substantial number of CC-licensed
open access articles from Wiley journals indexed in Crossref, likely
because of parsing errors on the publisher's website. This resulted in
fewer open access articles being recorded in Scopus and Web of Science.
Further inconsistencies emerged between Scopus and Web of Science,
despite both relying on Unpaywall: Web of Science erroneously labels
Elsevier's delayed open access as hybrid, whereas Scopus correctly
categorises it. Notably, Scopus and Crossref showed greater alignment
than that observed in a related comparison between Scopus and OpenAlex
(Alperin et al., 2024). This likely reflects the use of a curated list
to identify hybrid open access journals, rather than depending solely on
article-level tags. Given these variations, which may compromise
comparability, research and monitoring exercises should avoid relying on
a single source. Instead, the selection of open access evidence should
be cross-verified using multiple sources and snapshots that can be used
to track changes in the data over time (C.-K. Huang et al., 2020).
Incorporating expert-curated journal lists can also help prevent
misclassification based on business models (Visser et al., 2021).

It is important to note that the study's estimates of articles from
institutions involved in transformative agreements are approximations
because of a lack of access to invoice data, which is not usually shared
by library consortia and publishers. Furthermore, the open data sources
for transformative agreements, including the cOAlition S Journal Checker
Tool and the ESAC registry, are based on voluntary effort, crowd-sourced
from various consortia. However, recent validation using Dutch research
information demonstrates the reliability of such an open approach for
the assessment of articles under transformative agreements (de Jonge et
al., 2025). With enhanced open metadata compliance attributable to
evolving standards and initiatives in support of negotiations with
publishers, particularly through the ESAC initiative and Barcelona
Declaration on Open Research Information, the situation is likely to
improve. In conclusion, this study has shown that open metadata are
well-suited for analysing the transition from subscription-based journal
publishing to full open access. But using them in conjunction with
proprietary databases provides a more robust understanding that
overcomes individual shortcomings in coverage and metadata quality.

@@

@@

>>>>

These discussions highlight the need for robust data to evaluate
transformative agreements. Importantly, these agreements themselves can
help improve this evidence base, as researchers and libraries advocate
for including open metadata in negotiations to help avoid data analytics
from becoming a commodity controlled by publishers (Aspesi \& Brand,
2020; McCabe \& Mueller-Langer, 2024). For instance, the ESAC Workflow
Recommendations for Transformative and Open Access Agreements promote
using Crossref to share metadata that supports workflows including open
access licenses and funding information to support discovery and
monitoring (Geschuhn \& Stone, 2017). Simultaneously, the bibliometric
community pushes for more open metadata through this publisher-driven
DOI registration platform and continuously monitors progress (van Eck \&
Waltman, 2024).

@@

While promising for general bibliometric research, measuring hybrid open
accesss and transformative agreements presents particular challenges.
With invoicing data largely unavailable (Kramer, 2024), previous studies
have relied on publication fee pricing lists or expenditures, and first
and/or corresponding author affiliations as proxies to attribute funding
to institutions (Butler et al., 2023; Haustein et al., 2024; Zhang et
al., 2022). Others manually obtained journals and institutions from
transformative agreement documents before retrieving eligible
publications from the Web of Science (Bakker et al., 2024). Recent work
has extended this approach and programmatically attributed open access
articles to transformative agreements by using data from the cOAlition S
Journal Checker Tool, which links journals and participating
institutions to transformative agreements included in the ESAC registry,
and the open metadata sources Crossref and OpenAlex (de Jonge et al.,
2025; Jahn, 2025). Yet, no systematic comparison exists of how open
versus proprietary bibliometric data sources perform when analysing the
same hybrid journals covered by transformative agreements to identify
comparative strengths and shortcomings of data sources.

@@

hoaddata (Jahn, 2024), developed and maintained by the author to support
open access monitoring and research (Achterberg \& Jahn, 2023; Jahn,
2025), is an R data package that regularly collects information on
hybrid open access uptake from multiple openly available data sources.
It combines article-level metadata from Crossref (Hendricks et al.,
2020) and affilation metadata from OpenAlex (Priem et al., 2022) with
transformative agreement information from the Transformative Agreement
Data dump used by the cOAlition S Journal Checker Tool (JCT)\footnote{\url{https://journalcheckertool.org/transformative-agreements/}},
which links journal and institutional data about participating research
organisations to agreements in the ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}.

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
JCT journal data were enriched with ISSN variants linked to ISSN-L
according to the ISSN agency. To identify hybrid journals, a
comprehensive exclusion of fully open access journals was performed
using multiple journal lists including the Directory of Open Access
Journals (DOAJ). The resulting hybrid journal data comprising more than
13,000 hybrid journals were made available via hoaddata and were used to
independently determine the publication volume for each database.

@@

Figure \ref{fig-upset_coverage_results} presents the coverage of hybrid
journals included in transformative agreements, based on the
intersections of the examined data sources. The intersection sets of
journals (\ref{fig-upset_coverage_results}A) and articles
(\ref{fig-upset_coverage_results}B) are visualised as an UpSet graph
(Krassowski, 2020; Lex et al., 2014). The x-axis represents the set
intersections in a matrix layout. The analysis included hybrid journals
that published at least one open access article between 2019 and 2023,
based on open access status information from each database. Only
original articles were included in the analysis.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers, Elsevier,
Springer Nature, and Wiley, which collectively accounted for 47\% of the
hybrid journals and 62\% of the articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in 2,311
hybrid journals, although it recorded the largest number of hybrid
journals (19\%). Wiley accounted for 858,939 articles (12\%) in 1,382
hybrid journals (11\%). The remaining 54 publishers collectively
accounted for 2,749,847 articles (38\%) in 6,476 hybrid journals (53\%).

@@

Similarly, hybrid open access by transformative agreements substantially
increased between 2019 and 2023 (Figures \ref{fig-uptake_overview}B and
C). Trends were consistent when measuring first (dashed line) and
corresponding authorship (dotted line). According to Scopus, 479,297
open access articles could be attributed to transformative agreements
based on first author metadata (increasing from 15,341 to 206,084) and
489,262 using corresponding author metadata (from 15,444 to 210,816).
Web of Science recorded 493,028 open access articles via transformative
agreements using first author metadata (increasing from 21,871 to
174,126) and 500,076 using corresponding author metadata (from 22,092 to
177,030). hoaddata, lacking corresponding author data, linked 501,649
articles to transformative agreements, increasing from 16,010 in 2019 to
183,757 in 2023.

@@

When examining the percentage of open access articles in hybrid journals
(see Figure \ref{fig-uptake_country}B), a different pattern emerged.
Authors affiliated with institutions from medium-sized European
countries, such as Sweden, the Netherlands, Finland, and Hungary, made a
large proportion of their articles open access. Germany and the United
Kingdom also had approximately 40\% of their output available as open
access. In contrast, non-OECD countries showed notably lower adoption of
hybrid open access, with South Africa being the only BRICS member (as of
2022) well-represented in the data. The United States also demonstrated
a relatively low proportion of open access articles. These findings were
consistent across all databases. However, France was better represented
in Web of Science, likely because of its agreement with Elsevier
starting in 2019, which allowed delayed open access under the
publisher's user license (Rabesandratana, 2019). This licence was not
classified as hybrid open access in either Scopus or hoaddata. In all
cases, Spearman rank correlations were \(\rho > .9\), \(p < 0.001\),
indicating a high level of correlation between the databases and
authorship roles considered.

@@

The coverage analysis revealed that hybrid journals are well indexed in
all three data sources, particularly in terms of article coverage, which
reflects the dominance of major publishers whose established journal
portfolios are comprehensively indexed in proprietary databases (Bakker
et al., 2024; Bellen et al., 2024). Differences emerge for journals
targeting practitioners or local non-English language communities, with
many such titles indexed exclusively in Crossref and Scopus. Using
Crossref as a bibliometric database in hoaddata demonstrated a
particular strength in identifying newly established hybrid journals, a
notable finding given that transformative agreements primarily target
existing subscription-based journals. The landscape of hybrid journal
publishing thus differs markedly from that of fully open access
journals. Comparing the coverage of OpenAlex, Scopus and Web of Science,
Simard et al. (2024) indicate that only half of the fully open access
journals listed in the DOAJ are also indexed in Scopus and Web of
Science. Notably, journals that charge no publication fees (``diamond
journals'') are absent from the selective Web of Science, which
reinforces existing disparities in the indexing of under-represented
research communities and regions in selective bibliometric databases
(Simard et al., 2024).

@@

The database comparison revealed important discrepancies in open access
evidence. Crucially, not all publishers share CC licence metadata via
Crossref, a limitation that becomes apparent when contrasting Crossref
with Unpaywall's data in Scopus and Web of Science. While Unpaywall can
detect such gaps by parsing publishers' websites for open access
licences, it nevertheless missed a substantial number of CC-licensed
open access articles from Wiley journals indexed in Crossref, likely
because of parsing errors on the publisher's website. This resulted in
fewer open access articles being recorded in Scopus and Web of Science.
Further inconsistencies emerged between Scopus and Web of Science,
despite both relying on Unpaywall: Web of Science erroneously labels
Elsevier's delayed open access as hybrid, whereas Scopus correctly
categorises it. Notably, Scopus and hoaddata showed greater alignment
than that observed in a related comparison between Scopus and OpenAlex
(Alperin et al., 2024). This likely reflects the advantage of using of a
curated list to identify hybrid open access journals, rather than
depending solely on article-level tags. Given these variations, which
may compromise comparability, research and monitoring exercises should
avoid relying on a single source. Instead, the selection of open access
evidence should be cross-verified using multiple sources and snapshots
that can be used to track changes in the data over time (C.-K. Huang et
al., 2020). Incorporating expert-curated journal lists can also help
prevent misclassification based on business models (Visser et al.,
2021).

It is important to note that the study's estimates of articles from
institutions involved in transformative agreements are approximations
because of a lack of access to invoice data, which is not usually shared
by library consortia and publishers (Kramer, 2024). Furthermore, the
open data sources for transformative agreements, including the cOAlition
S Journal Checker Tool and the ESAC registry, are based on voluntary
effort, crowd-sourced from various consortia. However, recent validation
using Dutch research information demonstrates the reliability of such an
open approach for the assessment of articles under transformative
agreements (de Jonge et al., 2025). With enhanced open metadata
compliance attributable to evolving standards and initiatives in support
of negotiations with publishers, particularly through the ESAC
initiative and Barcelona Declaration on Open Research Information, the
situation is likely to improve. In conclusion, this study has shown that
open metadata are well-suited for analysing the transition from
subscription-based journal publishing to full open access. But using
them in conjunction with proprietary databases provides a more robust
understanding that overcomes individual shortcomings in coverage and
metadata quality.

@@

\bibitem[\citeproctext]{ref-Bakker_2024}
Bakker, C., Langham-Putrow, A., \& Riegelman, A. (2024). Impact of
transformative agreements on publication patterns: An analysis based on
agreements from the ESAC registry. \emph{International Journal of
Librarianship}, \emph{8}(4), 67--96.
\url{https://doi.org/10.23974/ijol.2024.vol8.4.341}

@@

\bibitem[\citeproctext]{ref-Kramer_2024}
Kramer, B. (2024). \emph{Study on scientific publishing in {Europe} --
{Development}, diversity, and transparency of costs}. Publications
Office of the European Union. \url{https://doi.org/doi/10.2777/89349}
